---
title: "Custom planning specs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom planning specs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Purpose

This vignette shows how to create your own `projflow` planning specs.

Use this workflow when the built-in components, deliverables, or presets are close to what you need, but not enough.

The three custom spec types are:

1. component specs: add new project capabilities such as a custom analysis step or governance block;
2. deliverable specs: add new output types such as a custom report or export bundle;
3. preset specs: save a reusable combination of components, deliverables, and infrastructure.

# Before you create a custom spec

Start by checking whether `projflow` already has what you need.

```{r}
library(projflow)

available_project_components()
available_project_deliverables()
available_project_presets()
```

Inspect the closest built-in option first.

```{r}
inspect_project_component("quality_control")
inspect_project_deliverable("client_report")
inspect_project_preset("client_report")
```

That gives you a good template before you create your own version.

It also shows an important design rule: dependencies are expanded automatically.

If a deliverable depends on a component, and that component depends on other components, `projflow` adds them during planning. If a preset defines a full workflow, `projflow` expands the preset exactly as defined.

# Option 1: Register a custom spec directly from R

The fastest way to experiment is to register a spec from a list.

## Custom component

```{r}
register_project_component(list(
  component = "genomic_qc",
  depends_on = "data_preparation",
  folders = c("outputs/genomics", "outputs/genomics/qc"),
  scripts = list(list(
    name = "genomic_qc",
    path = "analysis/08_genomic_qc.R",
    type = "quality_control",
    order = 80
  )),
  packages = c("dplyr", "readr"),
  checks = "genomic_qc_outputs_registered"
))
```

Inspect the result:

```{r}
inspect_project_component("genomic_qc")
```

## Custom deliverable

```{r}
register_project_deliverable(list(
  deliverable = "executive_summary",
  depends_on = "report",
  path = "reports/executive_summary.qmd",
  type = "report",
  packages = "quarto"
))
```

Inspect the result:

```{r}
inspect_project_deliverable("executive_summary")
```

The `depends_on` field is what tells `projflow` which components should be added automatically when a user requests the deliverable.

## Custom preset

```{r}
register_project_preset(list(
  preset = "rapid_review",
  components = c("data_preparation", "genomic_qc", "report"),
  deliverables = c("html_report", "executive_summary"),
  infrastructure = c("git"),
  description = "Compact preset for review-oriented analysis projects"
))
```

Inspect the result:

```{r}
inspect_project_preset("rapid_review")
```

A preset is different from a deliverable: it is already a complete named bundle of components, deliverables, and infrastructure.

# Option 2: Create YAML spec files

For reusable workflows, YAML files are usually better than inline R lists.

They can be stored in a repository, reviewed in pull requests, and shared across projects.

# Step 1: Create a component spec file

Create a file such as `genomic_qc.yml`.

```yaml
component: genomic_qc
depends_on:
  - data_preparation
folders:
  - outputs/genomics
  - outputs/genomics/qc
scripts:
  - name: genomic_qc
    path: analysis/08_genomic_qc.R
    type: quality_control
    order: 80
packages:
  - dplyr
  - readr
checks:
  - genomic_qc_outputs_registered
```

Load it into the current R session:

```{r}
use_project_component_spec("genomic_qc.yml")
inspect_project_component("genomic_qc")
```

# Step 2: Create a deliverable spec file

Create a file such as `executive_summary.yml`.

```yaml
deliverable: executive_summary
depends_on:
  - report
path: reports/executive_summary.qmd
type: report
packages:
  - quarto
```

Load it:

```{r}
use_project_deliverable_spec("executive_summary.yml")
inspect_project_deliverable("executive_summary")
```

# Step 3: Create a preset spec file

Create a file such as `rapid_review.yml`.

```yaml
preset: rapid_review
description: Compact preset for review-oriented analysis projects
components:
  - data_preparation
  - genomic_qc
  - report
deliverables:
  - html_report
  - executive_summary
infrastructure:
  - git
```

Load it:

```{r}
use_project_preset_spec("rapid_review.yml")
inspect_project_preset("rapid_review")
```

# Step 4: Confirm that the new options are available

Once a spec is registered, it becomes part of the same planning workflow as the built-in options.

```{r}
available_project_components()
available_project_deliverables()
available_project_presets()
```

# Step 5: Use the custom options in a project plan

After registering the specs, use them in `plan_project()` like any other option.

```{r}
plan <- plan_project(
  path = "custom-demo-project",
  components = c("data_preparation", "genomic_qc", "report"),
  deliverables = c("html_report", "executive_summary"),
  infrastructure = "git"
)

plan
plot(plan)
```

Or use the custom preset directly:

```{r}
plan <- plan_project(
  path = "custom-demo-project",
  preset = "rapid_review"
)

plan
plot(plan)
```

# How to decide which spec type to create

Create a custom component when you want to add:

1. folders;
2. starter scripts;
3. reports;
4. package suggestions;
5. checks;
6. component dependencies.

Create a custom deliverable when you want to add:

1. a new report path;
2. a new output folder;
3. a tracked file;
4. deliverable-specific dependencies on components.

Create a custom preset when you want to save:

1. a standard team workflow;
2. a project template you reuse often;
3. a named combination of components, deliverables, and infrastructure.

# Practical advice

Keep custom names short and stable. Prefer names like `genomic_qc`, `executive_summary`, and `rapid_review`.

Start from inspection, not invention. Inspect the closest built-in option first, then change only what you actually need.

When you are designing a reusable team workflow, prefer YAML spec files over inline R lists.

After registering a custom option, always inspect it and then build a small test plan before using it in a real project.
