---
title: "projflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{projflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

`projflow` helps you create and manage a reproducible R analysis project.

If you have never used the package before, the simplest way to think about it is:

1. plan the project;
2. create the project files;
3. register scripts, outputs, and reports;
4. check that the workflow is valid;
5. run the workflow and track the work around it.

The package is organised into four layers:

1. project structure;
2. executable analysis DAG;
3. governance;
4. interfaces and integrations.

The analysis layer is a Directed Acyclic Graph (DAG), where data inputs feed scripts, scripts produce outputs, outputs feed reports, and reports produce deliverables.

# Before you start

This vignette assumes that `projflow` is already installed.

We will create a small example project called `"demo-project"`.

```{r}
library(projflow)
```

# Step 1: Inspect the available scaffold options

Before choosing project settings, look at the built-in options that `projflow` already knows how to scaffold.

```{r}
available_project_components()
available_project_deliverables()
available_project_infrastructure()
available_project_presets()
```

For a first project, components and infrastructure are usually the most important choices:

1. `components` describe the kinds of analytical work the project will contain;
2. `deliverables` describe the final outputs you want to produce;
3. `infrastructure` describes technical support such as Git, `renv`, `targets`, or GitHub Actions;
4. `presets` provide pre-defined combinations when you want a faster starting point.

If you want to understand what a specific option will add before using it, inspect it directly:

```{r}
inspect_project_component("quality_control")
inspect_project_deliverable("client_report")
inspect_project_infrastructure("targets")
inspect_project_preset("client_report")
```

`projflow` expands dependencies automatically. This is important:

1. if you ask for a preset, `projflow` adds everything defined by that preset;
2. if you ask for a deliverable, `projflow` adds the components required for that deliverable;
3. if one of those components depends on other components, those are added as well.

For example, the `client_report` **preset** already includes a full workflow with components such as `data_preparation`, `quality_control`, `statistical_analysis`, `tables`, `figures`, `report`, and governance support.

```{r}
inspect_project_preset("client_report")
```

By contrast, the `client_report` **deliverable** adds only the components needed to produce that deliverable. You do not need to remember those dependencies yourself.

```{r}
deliverable_plan <- plan_project(
  path = "client-report-demo",
  deliverables = "client_report",
  infrastructure = "git"
)

deliverable_plan$components
deliverable_plan$deliverables
```

In other words, users should ask for the outcome they want and then inspect the resulting plan, rather than trying to manually add every dependency in advance. Add `components` only when the project needs work beyond the deliverable's required dependencies.

You can also create your own planning options when the built-in ones do not match your workflow.

For a complete guide to writing custom component, deliverable, and preset specs, see the separate vignette on custom planning specs.

Register a custom component directly from R:

```{r}
register_project_component(list(
  component = "custom_qc",
  depends_on = "data_preparation",
  folders = "outputs/custom_qc",
  scripts = list(list(
    name = "custom_qc",
    path = "analysis/08_custom_qc.R",
    type = "quality_control",
    order = 80
  )),
  packages = "dplyr"
))
```

Register a custom deliverable:

```{r}
register_project_deliverable(list(
  deliverable = "executive_summary",
  depends_on = "report",
  path = "reports/executive_summary.qmd",
  type = "report",
  packages = "quarto"
))
```

Register a custom preset:

```{r}
register_project_preset(list(
  preset = "rapid_review",
  components = c("data_preparation", "report"),
  deliverables = "html_report",
  infrastructure = "git",
  description = "Compact preset for quick review projects"
))
```

If you prefer YAML files, use the corresponding helpers:

```{r}
use_project_component_spec("custom_component.yml")
use_project_deliverable_spec("custom_deliverable.yml")
use_project_preset_spec("custom_preset.yml")
```

After registering custom options, they become available to the same discovery and inspection functions:

```{r}
available_project_components()
available_project_deliverables()
available_project_presets()

inspect_project_component("custom_qc")
inspect_project_deliverable("executive_summary")
inspect_project_preset("rapid_review")
```

# Step 2: Plan the project before creating files

Start by building a plan. This lets you inspect what `projflow` intends to create before anything is written to disk.

```{r}
plan <- plan_project(
  path = "demo-project",
  components = c("data_preparation", "statistical_analysis", "report"),
  infrastructure = c("git")
)
```

Inspect the plan:

```{r}
plan
plot(plan)
```

At this stage, no project folder has been created yet.

## A more complex plan

For a larger client-facing project, start from a preset and add the work areas that are specific to the analysis. Explicit components extend the preset, while deliverable and infrastructure dependencies are expanded automatically.

```{r}
complex_plan <- plan_project(
  path = "regulated-client-analysis",
  title = "Regulated client analysis",
  preset = "client_report",
  components = c(
    "exploratory_analysis",
    "model_diagnostics",
    "data_governance",
    "communication",
    "validation"
  ),
  deliverables = c("model_outputs", "status_report"),
  infrastructure = c(
    "github_actions",
    "targets",
    "tests",
    "lintr",
    "styler"
  ),
  use_internal_data_dirs = FALSE,
  include_example = TRUE
)

complex_plan
complex_plan$components
complex_plan$deliverables
complex_plan$infrastructure
plot(complex_plan)
```

This plan combines analytical work, client delivery, governance, data handling, validation, automated checks, and reproducibility support. Because `targets` depends on `renv` and GitHub Actions depends on Git, those infrastructure dependencies are included even though they were not repeated manually. The external-data setting remains explicit; set `use_internal_data_dirs = TRUE` only when project data should live inside the scaffold.

# Step 3: Create the project

Once the plan looks correct, create the project scaffold.

```{r}
new_project(plan = plan, open = FALSE)
```

This writes the initial folders, configuration files, and project registry.

# Step 4: Set up the project for the current session

Now initialise the project so `projflow` can check folders, read metadata, and report missing packages.

```{r}
setup <- setup_project("demo-project")
setup
```

If packages are missing, `setup_project()` tells you which ones are not installed.

# Step 5: Inspect the project structure

Before adding analysis objects, look at the scaffold that was created.

```{r}
project_layers()
project_structure("demo-project")
```

This is useful when you are still learning where scripts, reports, outputs, and metadata are stored.

# Step 6: Configure where your data lives

`projflow` is designed to keep raw or large data outside the repository when possible.

Set a local data root for your machine:

```{r}
set_project_data_root("C:/data/demo-project", root = "demo-project")
```

Build file paths under that data root:

```{r}
project_data_path("raw", "measurements.csv", root = "demo-project")
```

If you prefer to keep data inside the project, you can do that when you create the plan, but the external-data workflow is the recommended default.

# Step 7: Add analysis scripts

Create the scripts that will do the work. A beginner-friendly pattern is to start with one script for preparing data and one script for analysis.

```{r}
new_script(
  name = "import_and_clean_data",
  script_type = "data_preparation",
  root = "demo-project",
  open = FALSE
)

new_script(
  name = "fit_model",
  script_type = "statistical_analysis",
  root = "demo-project",
  open = FALSE
)
```

These functions create script files and register them in the project metadata. The generated scripts are placeholders: you still need to open them and write the actual analysis code.

# Step 8: Register outputs and reports

After you know what the scripts are supposed to produce, register those outputs explicitly.

```{r}
new_output(
  name = "analysis_dataset",
  type = "dataset",
  root = "demo-project"
)

new_table(
  name = "model_summary",
  root = "demo-project"
)

new_figure(
  name = "diagnostic_plot",
  root = "demo-project"
)

new_report(
  name = "main_report",
  root = "demo-project"
)
```

This keeps the analysis objects visible in the registry instead of leaving them implicit in filenames only.

# Step 9: Inspect and validate the analysis DAG

Once scripts and outputs are registered, inspect the project graph.

```{r}
dag <- project_analysis_dag("demo-project")
dag$nodes
dag$edges
```

Validate the graph before running anything:

```{r}
validate_project_dag(root = "demo-project")
topological_project_order("demo-project", type = "scripts")
```

If the DAG is valid, `projflow` can determine a safe execution order for the scripts.

# Step 10: Run the project

After you have added real analysis code to the generated scripts, run the workflow:

```{r}
run_project("demo-project")
```

`run_project()` prefers DAG order when the graph is valid.

# Step 11: Optionally generate a targets pipeline

If you want a `_targets.R` file based on the registered DAG, generate it from the project metadata.

```{r}
write_targets_pipeline("demo-project", overwrite = TRUE)
```

This is optional, but useful when you want to connect `projflow` to a `targets` workflow.

# Step 12: Track project management tasks

`projflow` can also track the work around the analysis without mixing that governance layer into the execution graph.

```{r}
add_project_task("Review model diagnostics", root = "demo-project")
add_project_risk("Input data may change during the project", root = "demo-project")
add_project_decision("Use REML for the mixed model", root = "demo-project")
add_project_milestone("Draft report complete", root = "demo-project")
```

Review the current project status:

```{r}
project_status_report(root = "demo-project")
```

# Step 13: Run checks and open the dashboard

When you want a higher-level review of the project, use the diagnostic helpers.

```{r}
check_project("demo-project", deep = TRUE)
diagnose_project("demo-project")
open_dashboard("demo-project")
```

# A practical first workflow

For a first project, the recommended order is:

1. create a plan with `plan_project()`;
2. create the scaffold with `new_project()`;
3. run `setup_project()`;
4. configure the data location with `set_project_data_root()`;
5. create scripts with `new_script()`;
6. register outputs and reports;
7. validate the DAG;
8. write the real analysis code in the generated scripts;
9. run the project;
10. use governance and diagnostics as the project grows.

That sequence is enough for a beginner to go from an empty folder to a structured reproducible analysis project.
