---
title: "Using projflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using projflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`projflow` creates analysis project scaffolds that are intentionally lightweight:

- code, reports, and small derived artefacts live in the repository;
- raw, large, and machine-specific data live outside the repository by default;
- a project registry links scripts, reports, and outputs;
- projects can be planned, built, checked, and extended incrementally.

The package is designed for analysis projects that need useful defaults without
forcing a large folder tree or unnecessary placeholder files.

# Core concepts

Most of the public API is easier to understand if you keep four concepts
separate:

- **component**: a project capability, such as `statistical_analysis`,
  `report`, `tables`, `figures`, `project_management`, or `shiny_app`;
- **deliverable**: an intended end product, such as `html_report`, `tables`,
  `dashboard`, or `status_report`;
- **infrastructure**: supporting tooling, such as `git`, `quarto`, `tests`,
  `github_actions`, or `renv`;
- **registry object**: a script, report, or output entry recorded in
  `.projflow/project_registry.yml`.

The distinction between **scripts**, **reports**, and **outputs** matters:

- a **script** is source code to execute;
- a **report** is a `qmd` or `Rmd` document to render;
- an **output** is an artefact created by a script, such as an `.rds`, `.csv`,
  `.png`, or rendered report file.

# Typical workflow

The normal lifecycle looks like this:

1. Inspect a scaffold with `plan_project()`.
2. Create a project with `new_project()`.
3. Configure external data roots with `set_project_data_root()`.
4. Add scripts and reports with `new_script()`, `new_report()`, `new_app()`, or
   `new_component()`.
5. Save artefacts explicitly with `save_project_object()`.
6. Run and render with `build_project()`.
7. Inspect project health with `check_project()` and `project_status()`.

```{r, eval = FALSE}
library(projflow)

plan <- plan_project(
  path = "my_analysis",
  components = c("data_preparation", "statistical_analysis", "report"),
  infrastructure = character()
)

new_project(
  path = "my_analysis",
  components = c("data_preparation", "statistical_analysis", "report"),
  infrastructure = character(),
  open = FALSE
)

set_project_data_root("D:/shared/my_analysis_data")

new_script("clean_phenotypes", script_type = "data_cleaning", open = FALSE)

build_project(render_reports = TRUE)
check_project()
```

# Planning a project before writing files

`plan_project()` previews what `new_project()` would create.

```{r, eval = FALSE}
plan <- plan_project(
  path = "demo-project",
  title = "Demo Project",
  components = c("statistical_analysis", "report"),
  infrastructure = c("git", "tests"),
  include_example = TRUE
)
```

The returned plan contains:

- `plan$components`, `plan$deliverables`, `plan$infrastructure`;
- `plan$folders` and `plan$files`;
- `plan$scripts` and `plan$reports`;
- `plan$registry`, which shows the script/report/output entries that will be
  written;
- `plan$checks`, which includes dependency-resolution messages.

This is especially useful when presets, deliverables, or dependencies would
expand the scaffold.

## Components, deliverables, and infrastructure

Use components to describe the analytical workflow:

```{r, eval = FALSE}
plan_project(
  path = "analysis-only",
  components = c("data_preparation", "quality_control", "statistical_analysis"),
  infrastructure = character()
)
```

Use deliverables when you care about outputs rather than only upstream steps:

```{r, eval = FALSE}
plan_project(
  path = "reporting-project",
  components = "statistical_analysis",
  deliverables = c("html_report", "tables"),
  infrastructure = character()
)
```

Use infrastructure to add reproducibility or automation support:

```{r, eval = FALSE}
plan_project(
  path = "tested-project",
  components = c("statistical_analysis", "report"),
  infrastructure = c("git", "tests", "github_actions")
)
```

## Presets and explicit arguments

If you use a preset, `projflow` merges it with explicit arguments rather than
replacing them.

```{r, eval = FALSE}
plan_project(
  path = "preset-project",
  preset = "basic_analysis",
  deliverables = "tables",
  infrastructure = c("git", "tests")
)
```

## Internal versus external data directories

By default, `projflow` assumes project data are external. If you explicitly want
internal `data/raw/` and `data/processed/` folders, set
`use_internal_data_dirs = TRUE`.

```{r, eval = FALSE}
new_project(
  path = "internal-data-project",
  components = c("statistical_analysis", "report"),
  use_internal_data_dirs = TRUE,
  infrastructure = character()
)
```

# Creating a new project

`new_project()` writes the scaffold to disk.

```{r, eval = FALSE}
new_project(
  path = "livestock-study",
  title = "Livestock Study",
  components = c("statistical_analysis", "report"),
  infrastructure = character(),
  include_example = TRUE,
  open = FALSE
)
```

For a minimal report-oriented project, the default scaffold usually includes:

- `README.md`;
- `run_project.R`;
- `project.yml`;
- `.projflow/project_registry.yml`;
- `.projflow/local.yml`;
- `analysis/`;
- `reports/`;
- `outputs/` with typed subdirectories (`data/`, `analysis/`, `tables/`, `figures/`, `models/`, `diagnostics/`, `reports/`);
- `<project>.Rproj`.

If `include_example = TRUE`, the scaffold also includes
`analysis/0_example_analysis.R` or another sequentially numbered script. This script is comment-only and does not create a placeholder output automatically.

## Minimal report scaffolds

The `report` component is intentionally minimal:

- it implies `html_report`;
- it creates `reports/main_report.qmd`;
- it does **not** imply `tables` or `figures`;
- it does **not** add a default summary script by itself.

That means you only get tables or figure scripts if you request those
components or deliverables explicitly.

# Configuring external data

External data roots are stored in `.projflow/local.yml`, which is intended
to stay machine-local and should not be committed.

## Set and read data roots

```{r, eval = FALSE}
set_project_data_root("/mnt/project_data", name = "default")
set_project_data_root("/mnt/reference_data", name = "reference")

project_data_root()
project_data_root("reference")
```

## Build data paths reproducibly

```{r, eval = FALSE}
phenotypes <- utils::read.csv(
  project_data_path("phenotypes.csv")
)

field_map <- project_data_path(
  "maps",
  "field_map.gpkg",
  source = "reference"
)
```

## Inspect and remove configured sources

```{r, eval = FALSE}
list_project_data_sources()
check_project_data_access()
remove_project_data_source("reference")
```

# Adding scripts, reports, apps, and components

## New scripts

`new_script()` is the main high-level helper for user-created scripts.

```{r, eval = FALSE}
new_script("clean_phenotypes", script_type = "data_cleaning", open = FALSE)
```

By default, this:

- creates a numbered script such as `analysis/1_clean_phenotypes.R`;
- registers the script in the project registry;
- does **not** register an output;
- does **not** create a placeholder `.rds`.

### Minimal script creation and separate output registration

`new_script()` deliberately keeps the user-facing API small. It always creates a documented, comment-only script and assigns the next sequential script number automatically.

```{r, eval = FALSE}
new_script(
  name = "merge_datasets",
  script_type = "data_preparation",
  open = FALSE
)
```

Expected outputs are registered separately after deciding what the script should deliberately produce.

```{r, eval = FALSE}
new_output(
  name = "heritability_model",
  type = "model",
  path = "outputs/models/heritability_model.rds"
)
```

### Full script constructor

`new_project_script()` exposes the same minimal behaviour directly:

```{r, eval = FALSE}
new_project_script(
  name = "quality_summary",
  script_type = "quality_control",
  open = FALSE
)
new_output(
  name = "qc_summary",
  type = "analysis",
  path = "outputs/analysis/qc_summary.rds"
)
```

## Convenience script helpers

These are wrappers around `new_project_script()` with preset types:

```{r, eval = FALSE}
new_project_model("genomic_model", open = FALSE)
new_project_table("export_tables", open = FALSE)
new_project_figure("plot_results", open = FALSE)
```

## Reports

Create or extend reports with `new_report()` or `new_project_report()`.

```{r, eval = FALSE}
new_report("main_report", type = "html_report", open = FALSE)
new_report("client_report", type = "client_report", open = FALSE)

new_project_report("supplementary_note", format = "qmd", open = FALSE)
```

`new_report()` is a higher-level verb that can add required components or
deliverables. `new_project_report()` just writes and registers a report file.

## Components

Add capabilities to an existing project with `new_component()`.

```{r, eval = FALSE}
new_component("tables", open = FALSE)
new_component("figures", open = FALSE)
new_component("project_management", open = FALSE)
```

## Apps and dashboards

```{r, eval = FALSE}
new_app(type = "shiny", open = FALSE)
new_app(type = "quarto_dashboard", open = FALSE)
```

`new_app()` adds the corresponding component or deliverable and writes the
expected entry-point file when needed.

# Outputs and the project registry

The registry is the backbone of the package.

Use these functions when you want explicit control over outputs.

## Register, update, and remove outputs

```{r, eval = FALSE}
register_project_object(
  name = "weekly_summary",
  path = "outputs/tables/weekly_summary.csv",
  type = "table"
)

update_project_object(
  "weekly_summary",
  generated_by = "summarise_weekly"
)

unregister_project_object("weekly_summary")
```

## Save and load outputs

`save_project_object()` writes a file to the appropriate output location and
records it in the registry if needed.

```{r, eval = FALSE}
result <- data.frame(id = 1:3, value = c(4.2, 4.8, 5.1))

save_project_object(
  result,
  name = "analysis_results",
  type = "analysis"
)

load_project_object("analysis_results")
```

Supported output families include:

- RDS-style data-preparation and dataset outputs under `outputs/data/`;
- RDS-style analysis and generic outputs under `outputs/analysis/`;
- model outputs under `outputs/models/`;
- model diagnostics under `outputs/diagnostics/`;
- `.csv` outputs for `table`, `summary`, and `export` under `outputs/tables/`;
- `.png` outputs for `figure` and `visualisation` under `outputs/figures/`;
- rendered reports under `outputs/reports/<report-name>/<report-name>.html`.

## Saving outside the repository

If you want to save derived artefacts under a configured external data root
instead of the project repository:

```{r, eval = FALSE}
save_external_project_object(
  result,
  name = "analysis_results_backup",
  type = "analysis",
  source = "default"
)
```

## Inspect registered objects and outputs

```{r, eval = FALSE}
list_project_objects()
list_project_outputs()
missing_project_outputs()
stale_project_outputs()
```

These helpers are useful for understanding what the registry expects, what has
actually been created, and which outputs are older than their generating
scripts.

# Running scripts and building the project

## Setup inside scripts

Analysis scripts should normally start with:

```{r, eval = FALSE}
projflow::setup_project()
```

This loads project metadata, locates the root, ensures key directories exist,
and reports missing package requirements.

## Run individual objects or steps

```{r, eval = FALSE}
run_project_object("analysis_results")
run_project_step("analysis_core")
```

## Run all registered scripts

```{r, eval = FALSE}
run_project()
```

`run_project()` executes registered scripts in order using the registry.

## Build the full project

```{r, eval = FALSE}
build_project(render_reports = FALSE)
build_project(render_reports = TRUE)
```

`build_project()` runs registered scripts and then optionally renders reports.

## Render only reports

```{r, eval = FALSE}
render_project_reports()
```

Quarto reports require Quarto to be installed and available. R Markdown reports
require `rmarkdown`.

# Serving reports and apps interactively

Use `serve_project()` when you want a lightweight preview entry point.

```{r, eval = FALSE}
serve_project(target = "project", render = FALSE)
serve_project(target = "reports", render = TRUE)
serve_project(target = "dashboard", render = TRUE)
serve_project(target = "shiny_app", render = FALSE)
```

With `target = "auto"`, `projflow` chooses a sensible default based on the
files present in the project.

# Checking project health

`check_project()` is a health check, not a full build.

```{r, eval = FALSE}
check_project()
check_project(deep = TRUE)
check_project(render_reports = TRUE)
check_project(strict = TRUE)
check_project(repair = TRUE)
```

Important behavior:

- `deep = TRUE` performs deeper static checks;
- `render_reports = TRUE` opts into rendering reports during the check;
- `strict = TRUE` raises an error on critical failures;
- `repair = TRUE` can recreate safe missing project files and directories.

Use `project_status()` for a lightweight status object with a dedicated print
method:

```{r, eval = FALSE}
status <- project_status()
status
```

## Git and GitHub Actions checks

```{r, eval = FALSE}
check_git_status()
check_github_actions()
use_github_actions(workflow = "check-project")
use_github_actions(workflow = "render-reports")
```

# Project management helpers

If you include the `project_management` component, `projflow` creates
`.projflow/tasks.yml` and governance files under `docs/`.

## Tasks

```{r, eval = FALSE}
add_project_task(
  "Review model outputs",
  status = "todo",
  priority = "high"
)

project_tasks()

update_project_task(
  "review_model_outputs",
  status = "blocked",
  notes = "Waiting for external data access"
)

complete_project_task("review_model_outputs")
remove_project_task("review_model_outputs")
```

## Milestones

```{r, eval = FALSE}
add_project_milestone("Draft report", due = "2026-06-01")
project_milestones()
update_project_milestone("draft_report", status = "done")
```

## Decisions and risks

```{r, eval = FALSE}
add_project_decision(
  title = "Use external data roots",
  decision = "Keep raw data outside the repository",
  rationale = "Avoid tracking large and machine-specific files"
)

project_decisions()

add_project_risk(
  title = "Missing external data",
  mitigation = "Confirm access before first analysis run"
)

project_risks()
```

## Status reports

```{r, eval = FALSE}
project_status_report(output = "data")
project_status_report(output = "markdown")
project_status_report(output = "html")
```

# Lower-level constructors

Most users will work with `new_project()`, `new_script()`, `new_report()`, and
`new_component()`, but the package also exports lower-level helpers:

```{r, eval = FALSE}
new_project_object("custom_table", type = "table")
new_project_idea("future_analysis", create_script = TRUE)
```

These are useful when you want direct control over exactly what gets written and
registered.

# End-to-end example

The following example shows a realistic minimal workflow.

```{r, eval = FALSE}
library(projflow)

new_project(
  path = "animal-breeding-study",
  components = c("data_preparation", "statistical_analysis", "report"),
  infrastructure = c("git", "tests"),
  include_example = TRUE,
  open = FALSE
)

set_project_data_root("D:/project_data/animal-breeding-study")

new_script(
  name = "clean_phenotypes",
  script_type = "data_cleaning",
  open = FALSE
)

new_script(
  name = "fit_heritability_model",
  script_type = "model",
  open = FALSE
)
new_output(
  name = "heritability_model",
  type = "model",
  path = "outputs/models/heritability_model.rds"
)

new_component("tables", open = FALSE)
new_component("figures", open = FALSE)

add_project_task("Check pedigree completeness", priority = "high")

build_project(render_reports = TRUE)

list_project_outputs()
project_status_report(output = "markdown")
check_project(deep = TRUE)
```

# Function map

The exported API can be grouped as follows.

## Project creation and planning

- `plan_project()`
- `new_project()`
- `new_component()`
- `new_report()`
- `new_app()`

## Data location helpers

- `set_project_data_root()`
- `project_data_root()`
- `project_data_path()`
- `list_project_data_sources()`
- `check_project_data_access()`
- `remove_project_data_source()`

## Scripts, reports, and outputs

- `new_script()`
- `new_project_script()`
- `new_project_report()`
- `new_project_model()`
- `new_project_table()`
- `new_project_figure()`
- `new_project_idea()`
- `new_project_object()`
- `register_project_object()`
- `update_project_object()`
- `unregister_project_object()`
- `save_project_object()`
- `save_external_project_object()`
- `load_project_object()`
- `list_project_objects()`
- `list_project_outputs()`
- `missing_project_outputs()`
- `stale_project_outputs()`

## Running and rendering

- `setup_project()`
- `run_project()`
- `run_project_object()`
- `run_project_step()`
- `build_project()`
- `render_project_reports()`
- `serve_project()`

## Checks and automation

- `check_project()`
- `project_status()`
- `check_git_status()`
- `check_github_actions()`
- `use_github_actions()`

## Project management

- `project_tasks()`
- `add_project_task()`
- `update_project_task()`
- `complete_project_task()`
- `remove_project_task()`
- `project_milestones()`
- `add_project_milestone()`
- `update_project_milestone()`
- `project_decisions()`
- `add_project_decision()`
- `project_risks()`
- `add_project_risk()`
- `project_status_report()`

# Summary

`projflow` is most effective when you use it as a thin project orchestration
layer:

- keep the scaffold minimal;
- keep raw and large data external by default;
- register scripts, reports, and outputs intentionally;
- save outputs explicitly rather than relying on placeholders;
- use `build_project()` for execution and `check_project()` for health checks.

That workflow keeps projects reproducible without forcing unnecessary structure
or repository clutter.
