---
title: "projflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{projflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

`projflow` creates and manages reproducible R analysis projects. The package is organised into four layers:

1. project structure;
2. executable analysis DAG;
3. governance;
4. interfaces and integrations.

The executable layer is a Directed Acyclic Graph: data inputs feed scripts, scripts produce outputs, outputs feed reports, and reports produce deliverables.

# Create a project

```{r}
library(projflow)

plan <- plan_project(
  path = "demo-project",
  components = c("data_preparation", "statistical_analysis", "report"),
  infrastructure = c("git", "targets")
)

new_project(plan = plan, open = FALSE)
setup_project("demo-project")
```

# Inspect the structure layer

```{r}
project_layers()
project_structure("demo-project")
```

# Register analysis objects

```{r}
new_script("clean_inputs", script_type = "data_preparation", root = "demo-project")
new_script("fit_model", script_type = "statistical_analysis", root = "demo-project")
new_table("model_summary", root = "demo-project")
new_figure("diagnostic_plot", root = "demo-project")
new_report("main_report", root = "demo-project")
```

# Work with the analysis DAG

```{r}
dag <- project_analysis_dag("demo-project")
dag$nodes
dag$edges

validate_project_dag(dag = dag)
topological_project_order("demo-project", type = "scripts")
```

`run_project()` uses the DAG order when the DAG is valid.

```{r}
run_project("demo-project")
```

# Generate a targets pipeline

```{r}
write_targets_pipeline("demo-project", overwrite = TRUE)
```

# Governance

```{r}
add_project_task("Review model diagnostics", root = "demo-project")
add_project_risk("External data dictionary may change", root = "demo-project")
add_project_decision("Use REML for the mixed model", root = "demo-project")
add_project_milestone("Draft report complete", root = "demo-project")

project_status_report(root = "demo-project")
```

# Interfaces and diagnostics

```{r}
check_project("demo-project", deep = TRUE)
diagnose_project("demo-project")
open_dashboard("demo-project")
```
